BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2016-01-13 Paper/BOAST.tex
Finished implementing Fred remarks.

\newcommand{\Specfem}{\productname{Specfem 3D}\xspace}
\newcommand{\Cuda}{\productname{Cuda}\xspace}
\newcommand{\OCL}{\productname{OpenCL}\xspace}
\newcommand{\code}[1]{\texttt{#1}}

@@

\paragraph{Temporary Variables Size} Using \textit{int} to store intermediary
results is unnecessary. Listing~\ref{lst:laplace_opencl_filter_short} show how
the code is modified to use smaller types. This yields a speedup of 1.3 for most
image sizes.

\lstset{style=customcl,
        caption={Laplace OpenCL Filter Vectorized Using \textit{short}.},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter_short}}
\begin{lstlisting}
  short16 v11 = convert_short16(v11_);
  short16 v12 = convert_short16(v12_);
  short16 v13 = convert_short16(v13_);
  short16 v21 = convert_short16(v21_);
  short16 v22 = convert_short16(v22_);
  short16 v23 = convert_short16(v23_);
  short16 v31 = convert_short16(v31_);
  short16 v32 = convert_short16(v32_);
  short16 v33 = convert_short16(v33_);

@@

In this section we will present the different aspects of BOAST run-time that
allow this automation. Those aspects include:
  \begin{itemize}
  \item multi target language generation ( performance, portability )
  \item compilation ( productivity, performance )
  \item execution ( productivity, performance )
  \item tracer, dumper and replay for non regression tests ( productivity )
  \end{itemize}

@@

Hence, the verification and validation of the new port can be
narrowed down to asserting that both codes apply the same operations
on the GPU. And the debugging part will consist in understanding what
diverges. To that purpose, we developed \code{GPUTrace}, inspired by
\code{strace} and \code{ltrace} tools: \code{GPUTrace} dynamically
preloads a library between the application and the GPU library, and
collects the function name, execution range and argument values (input
\emph{and} output) of the relevant function calls. These information
are traced in a unified format for all the APIs. This is a custom
trace format. Listing~\ref{lst:GPUTrace} presents a sample output
generated during \Specfem tracing.

@@

In this section we will present the benefits of using BOAST on the Laplace
motivating example as well as two scientific applications that use BOAST. The
first one, BigDFT~\cite{genovese2008daubechies}, uses BOAST in order to develop new
functionalities with performance portability in mind. The second one,
SPECFEM3D~\cite{komatitsch2011fluid}, uses BOAST to factorize OpenCL and CUDA
development, while having robust non regression tests.

@@

Here are the parameters we finally selected for our kernel optimization and
their possible values:
\begin{itemize}
\item \textit{x\_component\_number}: a positive integer
\item \textit{y\_component\_number}: a positive integer
\item \textit{vector\_length}: 1, 2, 4, 8 or 16
\item \textit{temporary\_size}: 2 or 4
\item \textit{synthesize\_loads}: \textit{true} or \textit{false}
\item \textit{vector\_recompute}: \textit{true} or \textit{false}
\end{itemize}
The last parameter is used when \textit{x\_component\_number} is not divisible by
\textit{vector\_length}. Two solutions are possible then, divide the remainder
of the division in vectors of smaller sizes (\textit{vector\_recompute} =
\textit{false}) or load more data and compute useless values in vectors of the
specified size (\textit{vector\_recompute} = \textit{true}).
This last option mimics the behavior of the ARM implementation, although when
working at the component level it may not be a valuable thing to try.

@@

Table~\ref{tbl:ARM} show the best results obtained by ARM on different images
compared to the naive implementation and to the best version version BOAST
found. It shows that the generated version systematically outperforms the hand
optimized version. As far as the optimization options are concerned, the results
are disappointing: the same kernel configuration is the best for all image
sizes. This kernel uses \textit{x\_component\_number} = 16,
\textit{y\_component\_number} = 1, \textit{vector\_length} = 16,
\textit{temporary\_size} = 2 and \textit{synthesize\_loads} = false.
\textit{vector\_recompute} because \textit{x\_component\_number}
== \textit{vector\_length}. Those results show that, when working on full
vectors, synthesizing the loads is harmful to performance and the programmer is
better of trusting the cache to load each vector in one cycle without
compromising the bandwidth. The results shown here are the best of four runs for
each configuration.

@@

The optimization parameters selected are not the same for those architectures.
Indeed, the Intel CPU favors kernels that have the parameters,
\textit{x\_compo-nent\_number} = 16, \textit{vector\_length} = 8,
\textit{temporary\_size} = 2 and \textit{synthesize\_loads} = false. Once again
\textit{vector\_recompute} does not apply. \textit{y\_component\_number} varies
from 4 to 2 when image size increase, thus decreasing task granularity as the
global work size increase. This result is unexpected and understanding why this
happen could be interesting. The NVIDIA GPU favors processing square tiles:
\textit{x\_component\_number} = 4, \textit{y\_component\_number} = 4,
\textit{vector\_length} = 4, \textit{temporary\_size} = 2 and
\textit{synthesize\_loads} = false. Once more, \textit{vector\_recompute} has no
meaning.

@@

The number of convolution kernels needed in BigDFT has been continuously
growing in the recent years.  Various boundary conditions and functionalities
have been added, making the BigDFT more and more powerful in terms of
scientific applications.  However, the cost of the maintenance and of
development of the convolutions is always a delicate point to be considered
while including a new functionality.  The convolution patterns are usually
rather similar, leading to code duplication and difficulties in code
maintainability.

It appears therefore very interesting to benefit from an automatic tool to drive
the implementation and the generation of new convolutions.  This would lead to
an optimized code, adapted to different computing platforms, that is optimally
factorized.  In addition to this point, the help of such code generator is also
important to build \emph{new science}: the cost of implementing new
convolutions would become so little that other functionalities (for example,
the generalization of the BigDFT convolutions to Neumann boundary conditions
or the usage of wavelet-on-the-interval basis) can be added with limited
manpower.

@@

The interest in having a robust and optimally tuned library in this scientific
field is therefore evident. Techniques are under investigation to provide also
fine-tuned binaries to the end-user rather than the code sources, such that
more aggressive inter procedural optimization can be performed. Indeed, BOAST
finds the optimal source code for a given kernel and compiler configuration but
one could imagine using binary optimizer to the compiled binaries. Those
optimizers could be coupled to BOAST and the output of the process would be the
final binary rather than the source code. This experience therefore would be a
first step toward the release of a tunable optimised convolution library
oriented to computational physics communities.

    \subsubsection{Performance report}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Res_synthesis}
\end{center}
\caption{Impact of Unrolling, Language and OpenMP on a Wavelet Transform}
\label{fig:synthesis}
\end{figure}

@@

In this last subsection, we study a free seismic wave propagation
simulator, \Specfem, in its \productname{Globe}
flavor\footnote{Specfem3d Globe --- CIG
  \url{http://www.geodynamics.org/cig/software/specfem3d-globe}}. \Specfem
simulates seismic wave propagation at the local or regional scale
based upon spectral-element method (SEM), with very good accuracy and
convergence properties. It is a reference application for super
computer benchmarking, thanks to its good scaling capabilities.

@@

When we started to work on the project (version v2.1 of July 2013) it
supported graphics card GPU acceleration through NVidia \Cuda. This
GPU support came in addition to the MPI support implemented to enable
multi-CPU parallel computing. Most of \Specfem code-base is written in
\productname{Fortran2003}, only the GPU related parts are written in
C.  The split between CPU and GPU code was done at a rather fine
grain, as the application counted more than 40 GPU kernels. Some of
them were quite simple (\eg performing a few vector operations, but at
massively parallel scale), while at the other side of the spectrum,
some complex kernels took more than 80 parameters and perform very
specific physical transformations.

@@

As part of the Mont-Blanc project, we had to port \Specfem to \OCL, so
that it could be used to benchmark Mont-Blanc HPC platforms.

@@

Nvidia \Cuda and \OCL are based on the same programming model: a
massively parallel accelerator running in disjoint, non addressable
memory environment. Thanks to that proximity, we have been able to carry
out most of the porting task with only a limited knowledge about
\Specfem internal physics.

This lack of \Specfem internal knowledge led us to be particularly
careful to the path we undertook for the port, as we would have
been unable to understand how and why the application was not
operating properly, if it was to fail.

Hence, our first milestone in the port was the translation of
\Specfem's \Cuda kernels into BOAST EDSL. This way, we could ask BOAST
framework to generate a CUDA version of the kernels, plug them back
into \Specfem and get (after fixing compile-time errors---prototypes
and naming mistakes mainly) a first set of \Specfem seismograms.

As we had expected, the seismograms were erroneous. But with the help
of shell scripts and BOAST framework ability to store and provide the
kernels' original source code, we built a set of \Specfem binaries
including only \emph{one} BOAST-generated kernel, with all others
reference-kernels. Running and validating all these binaries enabled
use to pinpoint the misbehaving kernels. We finished the debugging
with a side-by-side comparison that highlighted the coding mistakes.

\subparagraph{Porting run-time to \OCL} The second part of the port
consisted in the translation of the CPU-side of the application, from
\Cuda API to \OCL API. Most of the functions of the interfaces are
very similar, with only naming-convention and data-structure
distinctions. Hence, it was clear that automatic rewriting tools
(namely \code{sed} regexp and \code{emacs-lisp} functions) could be
useful. To give an idea of the cost of a \emph{manual} rewriting, we
can count (in \# of \OCL API function calls): 70 kernel ``function
calls'', 790 arguments to set, 230 memory transfers, 160 buffer
creations, and 270 releases.

Once the transformations were applied,
compilation errors fixed, and \OCL unsuccessful function calls solved,
the application managed to complete its execution and generate a first
set of seismograms. And again, as expected and feared, these
seismograms were not valid as their shape was completely different from the
reference ones.

As we had already validated BOAST-generated kernels (and trusted \Cuda and \OCL
versions to be semantically identical), we knew that the bugs were now in the
usage of the run-time, and we had to find a way to understand where \Specfem's
\Cuda version of the code diverged from its \OCL counterpart. To help us in
that purpose, we had a strong assumption: both versions of the code were
supposed to perform exactly the same operations, with the same ``logical''
parameters (the APIs have \emph{implementation} differences, for instance \OCL
has two memory transfer functions, \code{clEnqueueReadBuffer},
\code{clEnqueueWriteBuffer}, whereas \Cuda has only one, with a direction
parameter \code{cudaMemcpy(..., dir)}, but above that, it is the same
functionality).

Hence, our idea for locating the execution problems was to make sure
that both execution actually did the same thing. As the \OCL results
were invalid, we knew the executions would diverge at one or several
points.

\subparagraph{Debugging \OCL Execution: \code{GPUTrace}}

With the help of \code{GPUTrace} (Section~\ref{sec:gputrace}), we
could confront \Cuda and \OCL execution traces with a graphical
\code{diff} tool, and spot the different porting mistakes: some
parameters reversed, offsets incorrectly applied, \etc{}.

One last problem remained, clearly highlighted by the seismograms not
matching perfectly (they had a similar shape, but with a reduced
intensity). We added more verbosity to \code{GPUTrace} output: first
the initial bits of the GPU memory buffers, then their full
content. The drift was visible in the trace, but it was nonetheless
unclear where it started. We finally got it after hours of code review
of BOAST kernels and \OCL code. One kernel was
\emph{three}-dimensional, whereas the others were two-dimensional. But
for all of them, only two dimensions were passed, and one was
missing.
%As it was the same person who coded the \OCL port and the GPU
%tracer, this third dimension was forgotten in both codes.

\subparagraph{Evaluation} Our \OCL/BOAST port of \Specfem is now
merged in \Specfem's development tree and under test and extension by
different research teams. On a platform with two K40x GPUs and N Intel
Xeon processors, we measured similar results between the original
\Cuda version and our BOAST-\Cuda version. With the same set of
optimization flags, BOAST \Cuda and \OCL version reported similar
execution time spans. The best execution speed was achieved with \Cuda
version though (25\% higher than \OCL), as one optimization parameter
(\code{CUDA\_LAUNCH\_BOUND}) cannot be passed, to \OCL run-time, as of version
1.1. This parameter, in addition to specifying the work group size (which can
be done in \OCL), also constrains the number of work group that must run in
parallel on a multiprocessor. This value is set to 7 in \Cuda. This means that
the compiler must be very conservative on register usage in order to allow this
parallelism which allows better overlapping of communications end computations.
This functionality is not supported in \OCL.

By refactoring GPU kernel code in BOAST EDSL, the size of kernel code
shrank by a factor of 1.8 (from 7500 to less than 4000 LOC, partly
because of code duplication, also removing manually unrolled
loops). This is beneficial for \Specfem as it improves the readability
and maintainability of its source code.

We have also been able to enhance \Specfem's non-regression test-suite
by adding per-kernel non-regression tests. This was done with the help
of \code{GPUTrace}, that we used to capture all the input parameters
of a particular \emph{valid} kernel execution, as well as the output
values. Then, during the non-regression testing, BOAST framework
loads these buffer files, allocates GPU memory and initializes it through
\Cuda or \OCL run-time, and triggers the kernel execution. A comparison
of the output values (for instance against a maximal error level)
validates the non-regression.

In the same mindset, we provided \Specfem test-suite with kernel
performance evaluation mechanisms. These tests will help developers to
try new optimizations in kernel code and measure their impact, without
executing the whole application.

@@

  The idea to describe computing kernels using a DSL has been already explored.
SPIRAL~\cite{puschel2004spiral} is a decade old generation
framework for signal processing. It uses a proprietary DSL, SPL (Signal
Processing Language), to describe a DSP algorithm. This DSL is then transformed
into efficient programs in high level languages like C or Fortran.
POET~\cite{yi2007poet} also uses a DSL to describe custom code transformations,
like loop unrolling, loop blocking and loop interchange. Those transformation
can be parametrized in order to tune the application. Orio~\cite{Hart2009:Orio}
can be compared to POET as it aims at describing possible code transformations
using a DSL. All those approaches are very different from our as they put the
emphasis on compilation techniques whereas BOAST relies on the user to express
the different optimizations.

@@

%\section*{References}

>>>>

\newcommand{\code}[1]{\texttt{#1}}

@@

\paragraph{Temporary Variables Size} Using \emph{int} to store intermediary
results is unnecessary. Listing~\ref{lst:laplace_opencl_filter_short} show how
the code is modified to use smaller types. This yields a speedup of 1.3 for most
image sizes.

\lstset{style=customcl,
        caption={Laplace OpenCL Filter Vectorized Using \emph{short}.},
        breaklines=true,
        breakautoindent=false,
        label={lst:laplace_opencl_filter_short}}
\begin{lstlisting}
  short16 v11 = convert_short16(v11_);
  short16 v12 = convert_short16(v12_);
  short16 v13 = convert_short16(v13_);
  short16 v21 = convert_short16(v21_);
  short16 v22 = convert_short16(v22_);
  short16 v23 = convert_short16(v23_);
  short16 v31 = convert_short16(v31_);
  short16 v32 = convert_short16(v32_);
  short16 v33 = convert_short16(v33_);

@@

In this section we will present the different aspects of BOAST run-time that
allow this automation. Those aspects are multi-target language generation
(\emph{performance}, \emph{portability}), kernel compilation
(\emph{productivity}, \emph{performance}), kernel execution
(\emph{productivity}, \emph{performance}) and last, kernel tracer, dumper and
replay for non regression tests (\emph{productivity}).

@@

Hence, the verification and validation of the new port can be
narrowed down to asserting that both codes apply the same operations
on the GPU. And the debugging part will consist in understanding what
diverges. To that purpose, we developed \code{GPUTrace}, inspired by
\code{strace} and \code{ltrace} tools: \code{GPUTrace} dynamically
preloads a library between the application and the GPU library, and
collects the function name, execution range and argument values (input
\emph{and} output) of the relevant function calls. These information
are traced in a unified format for all the APIs. This is a custom
trace format. Listing~\ref{lst:GPUTrace} presents a sample output
generated during SPECFEM3D tracing.

@@

In this section we will present the benefits of using BOAST on the Laplace
motivating example as well as on two scientific applications. The first one,
BigDFT~\cite{genovese2008daubechies}, uses BOAST in order to develop new
functionalities with performance portability in mind. The second one,
SPECFEM3D~\cite{komatitsch2011fluid}, uses BOAST to factorize OpenCL and CUDA
development, while having robust non regression tests.

@@

Here are the parameters we finally selected for our kernel optimization and
their possible values:
\begin{itemize}
\item \emph{x\_component\_number}: a positive integer
\item \emph{y\_component\_number}: a positive integer
\item \emph{vector\_length}: 1, 2, 4, 8 or 16
\item \emph{temporary\_size}: 2 or 4
\item \emph{synthesize\_loads}: \emph{true} or \emph{false}
\item \emph{vector\_recompute}: \emph{true} or \emph{false}
\end{itemize}
The last parameter is used when \emph{x\_component\_number} is not divisible by
\emph{vector\_length}. Two solutions are possible then, divide the remainder
of the division in vectors of smaller sizes (\emph{vector\_recompute} =
\emph{false}) or load more data and compute useless values in vectors of the
specified size (\emph{vector\_recompute} = \emph{true}).
This last option mimics the behavior of the ARM implementation, although when
working at the component level it may not be a valuable thing to try.

@@

Table~\ref{tbl:ARM} show the best results obtained by ARM on different images
compared to the naive implementation and to the best version version BOAST
found. It shows that the generated version systematically outperforms the hand
optimized version. As far as the optimization options are concerned, the results
are disappointing: the same kernel configuration is the best for all image
sizes. This kernel uses \emph{x\_component\_number} = 16,
\emph{y\_component\_number} = 1, \emph{vector\_length} = 16,
\emph{temporary\_size} = 2 and \emph{synthesize\_loads} = false.
\emph{vector\_recompute} because \emph{x\_component\_number}
== \emph{vector\_length}. Those results show that, when working on full
vectors, synthesizing the loads is harmful to performance and the programmer is
better of trusting the cache to load each vector in one cycle without
compromising the bandwidth. The results shown here are the best of four runs for
each configuration.

@@

The optimization parameters selected are not the same for those architectures.
Indeed, the Intel CPU favors kernels that have the parameters,
\emph{x\_compo-nent\_number} = 16, \emph{vector\_length} = 8,
\emph{temporary\_size} = 2 and \emph{synthesize\_loads} = false. Once again
\emph{vector\_recompute} does not apply. \emph{y\_component\_number} varies
from 4 to 2 when image size increase, thus decreasing task granularity as the
global work size increase. This result is unexpected and understanding why this
happen could be interesting. The NVIDIA GPU favors processing square tiles:
\emph{x\_component\_number} = 4, \emph{y\_component\_number} = 4,
\emph{vector\_length} = 4, \emph{temporary\_size} = 2 and
\emph{synthesize\_loads} = false. Once more, \emph{vector\_recompute} has no
meaning.

@@

The number of convolution kernels needed in BigDFT has been continuously
growing in the recent years.  Various boundary conditions and functionalities
have been added, making the BigDFT more and more powerful in terms of
scientific applications.  However, the cost of maintenance and development of
the convolutions is always a delicate point to be considered while including a
new functionality.  The convolution patterns are usually rather similar,
leading to code duplication and difficulties in code maintainability.

Therefore it appears very interesting to benefit from an automatic tool to drive
the implementation and the generation of new convolutions.  This would lead to
an optimized code, adapted to different computing platforms, that is optimally
factorized.  In addition to this point, the help of such code generator is also
important to build \emph{new science}: the cost of implementing new
convolutions would become so little that other functionalities (for example,
the generalization of the BigDFT convolutions to Neumann boundary conditions
or the usage of wavelet-on-the-interval basis) can be added with limited
manpower.

@@

The interest in having a robust and optimally tuned library in this scientific
field is therefore evident. Techniques are under investigation to also provide
end users with fine-tuned binaries rather than the source codes, such that more
aggressive inter procedural optimization can be performed. Indeed, BOAST finds
the optimal source code for a given kernel and compiler configuration but one
could imagine using binary optimizer to the compiled binaries. Those optimizers
could be coupled to BOAST and the output of the process would be the final
binary rather than the source code. This experience therefore would be a first
step toward the release of a tunable optimized convolution library oriented to
computational physics communities.

    \subsubsection{Performance Report}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\linewidth]{Res_synthesis}
\end{center}
\caption{Impact of Unrolling, Language Used and OpenMP on a Wavelet Transform Code.}
\label{fig:synthesis}
\end{figure}

@@

In this last subsection, we study a free seismic wave propagation simulator,
SPECFEM3D, in its GLOBE flavor\footnote{Specfem3d Globe --- CIG
\url{http://www.geodynamics.org/cig/software/specfem3d-globe}}. SPECFEM3D
simulates seismic wave propagation at the local or regional scale based upon
spectral-element method (SEM), with very good accuracy and convergence
properties. It is a reference application for supercomputer benchmarking,
thanks to its good scaling capabilities.

@@

When we started to work on the project (version v2.1 of July 2013) it
supported graphics card GPU acceleration through NVidia CUDA. This
GPU support came in addition to the MPI support implemented to enable
multi-CPU parallel computing. Most of SPECFEM3D code-base is written in
Fortran 2003 and only the GPU related parts are written in
C.  The split between CPU and GPU code was done at a rather fine
grain, as the application counted more than 40 GPU kernels. Some of
them were quite simple (\eg performing a few vector operations, but at
massively parallel scale), while at the other side of the spectrum,
some complex kernels took more than 80 parameters and perform very
specific physical transformations.

@@

As part of the Mont-Blanc project, we had to port SPECFEM3D to OpenCL, so
that it could be used to benchmark Mont-Blanc HPC platforms.

@@

Nvidia CUDA and OpenCL are based on the same programming model: a
massively parallel accelerator running in disjoint, non addressable
memory environment. Thanks to that proximity, we have been able to carry
out most of the porting task with only a limited knowledge about
SPECFEM3D internal physics.

This lack of SPECFEM3D internal knowledge led us to be particularly
careful to the path we undertook for the port, as we would have
been unable to understand how and why the application was not
operating properly, if it was to fail.

Hence, our first milestone in the port was the translation of
SPECFEM3D's CUDA kernels into BOAST EDSL. This way, we could ask BOAST
framework to generate a CUDA version of the kernels, plug them back
into SPECFEM3D and get (after fixing compile-time errors---prototypes
and naming mistakes mainly) a first set of SPECFEM3D seismograms.

As we had expected, the seismograms were erroneous. But with the help
of shell scripts and BOAST framework ability to store and provide the
kernels' original source code, we built a set of SPECFEM3D binaries
including only \emph{one} BOAST-generated kernel, with all others
reference-kernels. Running and validating all these binaries enabled
use to pinpoint the misbehaving kernels. We finished the debugging
with a side-by-side comparison that highlighted the coding mistakes.

\subparagraph{Porting run-time to OpenCL} The second part of the port
consisted in the translation of the CPU-side of the application, from
CUDA API to OpenCL API. Most of the functions of the interfaces are
very similar, with only naming-convention and data-structure
distinctions. Hence, it was clear that automatic rewriting tools
(namely \code{sed} regexp and \code{emacs-lisp} functions) could be
useful. To give an idea of the cost of a \emph{manual} rewriting, we
can count (in \# of OpenCL API function calls): 70 kernel ``function
calls'', 790 arguments to set, 230 memory transfers, 160 buffer
creations, and 270 releases.

Once the transformations were applied,
compilation errors fixed, and OpenCL unsuccessful function calls solved,
the application managed to complete its execution and generate a first
set of seismograms. And again, as expected and feared, these
seismograms were not valid as their shape was completely different from the
reference ones.

As we had already validated BOAST-generated kernels (and trusted CUDA and OpenCL
versions to be semantically identical), we knew that the bugs were now in the
usage of the run-time, and we had to find a way to understand where SPECFEM3D's
CUDA version of the code diverged from its OpenCL counterpart. To help us in
that purpose, we had a strong assumption: both versions of the code were
supposed to perform exactly the same operations, with the same ``logical''
parameters (the APIs have \emph{implementation} differences, for instance OpenCL
has two memory transfer functions, \code{clEnqueueReadBuffer},
\code{clEnqueueWriteBuffer}, whereas CUDA has only one, with a direction
parameter \code{cudaMemcpy(..., dir)}, but above that, it is the same
functionality).

Hence, our idea for locating the execution problems was to make sure
that both execution actually did the same thing. As the OpenCL results
were invalid, we knew the executions would diverge at one or several
points.

\subparagraph{Debugging OpenCL Execution: \code{GPUTrace}}

With the help of \code{GPUTrace} (described in Section~\ref{sec:gputrace}), we
could confront CUDA and OpenCL execution traces with a graphical \code{diff}
tool, and spot the different porting mistakes: some parameters reversed,
offsets incorrectly applied, \etc{}.

One last problem remained, clearly highlighted by the seismograms not
matching perfectly (they had a similar shape, but with a reduced
intensity). We added more verbosity to \code{GPUTrace} output: first
the initial bits of the GPU memory buffers, then their full
content. The drift was visible in the trace, but it was nonetheless
unclear where it started. We finally got it after hours of code review
of BOAST kernels and OpenCL code. One kernel was
\emph{three}-dimensional, whereas the others were two-dimensional. But
for all of them, only two dimensions were passed, and one was
missing.
%As it was the same person who coded the OpenCL port and the GPU
%tracer, this third dimension was forgotten in both codes.

\subparagraph{Evaluation} Our OpenCL/BOAST port of SPECFEM3D is now
merged in SPECFEM3D's development tree and under test and extension by
different research teams. On a platform with two K40x GPUs and N Intel
Xeon processors, we measured similar results between the original
CUDA version and our BOAST-CUDA version. With the same set of
optimization flags, BOAST CUDA and OpenCL version reported similar
execution time spans. The best execution speed was achieved with CUDA
version though (25\% higher than OpenCL), as one optimization parameter
(\code{CUDA\_LAUNCH\_BOUND}) cannot be passed to OpenCL run-time, as of version
1.1. This parameter, in addition to specifying the work group size (which can
be done in OpenCL), also constrains the number of work group that must run in
parallel on a multiprocessor. This value is set to 7 in CUDA. This means that
the compiler must be very conservative on register usage in order to allow this
parallelism which allows better overlapping of communications end computations.
This functionality is not supported in OpenCL.

By refactoring GPU kernel code in BOAST EDSL, the size of kernel code
shrank by a factor of 1.8 (from 7500 to less than 4000 LOC, partly
because of code duplication, also removing manually unrolled
loops). This is beneficial for SPECFEM3D as it improves the readability
and maintainability of its source code.

We have also been able to enhance SPECFEM3D's non-regression test-suite
by adding per-kernel non-regression tests. This was done with the help
of \code{GPUTrace}, that we used to capture all the input parameters
of a particular \emph{valid} kernel execution, as well as the output
values. Then, during the non-regression testing, BOAST framework
loads these buffer files, allocates GPU memory and initializes it through
CUDA or OpenCL run-time, and triggers the kernel execution. A comparison
of the output values (for instance against a maximal error level)
validates the non-regression.

In the same mindset, we provided SPECFEM3D test-suite with kernel
performance evaluation mechanisms. These tests will help developers to
try new optimizations in kernel code and measure their impact, without
executing the whole application.

@@

  The idea to describe computing kernels using a DSL has been already explored.
SPIRAL~\cite{puschel2004spiral} is a decade old generation
framework for signal processing. It uses a proprietary DSL called SPL (Signal
Processing Language), to describe a DSP algorithm. This DSL is then transformed
into efficient programs in high level languages like C or Fortran.
POET~\cite{yi2007poet} also uses a DSL to describe custom code transformations,
like loop unrolling, loop blocking and loop interchange. Those transformations
can be parametrized in order to tune the application. Orio~\cite{Hart2009:Orio}
can be compared to POET as it aims at describing possible code transformations
using a DSL. All those approaches are very different from ours as they put the
emphasis on compilation techniques whereas BOAST relies on the user to express
the different optimizations.

@@

\section*{References}
